# Reproducible Research: Peer Assessment 1

## Vasudev Daruvuri

## Loading and preprocessing the data
```{r}
#Unzip and load data as CSV
setwd('F:/Analytics/Moocs/DS-Coursera/datasciencecoursera/ReproducibleResearch/RepData_PeerAssessment1');

data <- read.csv(unzip("activity.zip", "activity.csv"))

```


## What is mean total number of steps taken per day?
```{r}
#Sum steps by day, create Histogram, and calculate mean and median. 
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

The mean is `r rmean` and the median is `r rmedian`.


## What is the average daily activity pattern?

```{r}
#Calculate average steps for each interval for all days, create Plot, and find interval with most average steps. 
steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

## Imputing missing values
```{r}
#Find all missing cases and impute missing interval values with average of the interval across all days.  
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
#Impute day 1 steps to 0.
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0

#Recount total steps by day and create Histogram. 
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Calculate new mean and median for imputed data. 
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)

#Calculate difference between imputed and non-imputed data.
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

#Calculate total difference.
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)

```
Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA. 

Zeroes were imputed for 10-01-2012 because it was the first day and would have been over 9,000 steps higher than the following day, which had only 126 steps. NAs then were assumed to be zeros to fit the rising trend of the data. 

The imputed data mean is `r rmean.i`
The imputed data median is `r rmedian.i`
The difference between the non-imputed mean and imputed mean is `r mean_diff`
The difference between the non-imputed mean and imputed mean is `r med_diff`
The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?

``` {r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```

